---
title: "Homework 6 Q1Q2"
author: "Margaret Swift"
date: '`r Sys.Date()`'
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=F}
#----------------------------------------------------------------------------
# LOADING
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, cache=TRUE)
pacman::p_load(ggplot2, rjags)
source('../clarkfunctions2020.R')
Rcpp::sourceCpp('../cppFns.cpp')
```

## Exercise 1 {#q1}
_Interpret the main effects and interactions in this model._

I've renamed the parameters used in the JAGS model to make it a bit easier to interpret. 

```{r q1hide, results='hide'}
# Here is a data set:
data   <- read.table('../dataFiles/FACEtrees.txt',header=T)
data$y <- log(data[,'dnow']) # growth rate -- log of diameter now
tmp    <- model.frame(y ~ j + i + diam*trt, data)
ij     <- apply( cbind(tmp$j, tmp$i), 1, paste0, collapse='-') # tree, plot
X      <- model.matrix(y ~ diam*trt, tmp)
y      <- tmp$y
n      <- length(y)

# Here is model object for jags, which is written to a file `growthFixed.txt`:
cat( "model{
    for(i in 1:n){
      y[i]  ~ dnorm(mu[i],tau)
      mu[i] <- intercept + diameter*X[i,2] + co2treat*X[i,3] + inax*X[i,4]
    }

    tau     ~ dgamma(0.001,0.001)
    sigma   <- sqrt(1/tau)
    
    intercept ~ dnorm(0.0,1.0E-06)
    diameter ~ dnorm(0.0,1.0E-06)
    co2treat ~ dnorm(0.0,1.0E-06)
    inax ~ dnorm(0.0,1.0E-06)
  }
    ", fill=T, file="growthFixed.txt"
)

# Here I run the model and look at the output:
treeData   <- list(y = y, X = X, n = n )
varNames <- c('intercept', 'diameter', 'co2treat', 'inax')
parNames <- c(varNames, 'sigma')
parInit <- function(){ list(sigma = runif(1,0,1)) }
growFit <- invisible(jags.model(data=treeData, file="growthFixed.txt"))
update(growFit)
growFit <- coda.samples(growFit, variable.names=varNames, n.iter=5000)

```

```{r q1plot}
tmp <- summary(growFit)
betaFixed <- cbind( tmp$statistics[,1:2], tmp$quantiles[,c('2.5%','97.5%')] )
print(betaFixed)
par(mar=c(3,3,1,1))
plot( as.mcmc(growFit) )
```

## Exercise 2 {#q2}

_Run the model again, including `nfert` (nitrogen fertilization) in addition to other variables.  Interpret its effect._

Results from the model below. With the addition of the `nfert` predictor, the density outputs from JAGS remain the same for the CO$_2$ treatment, tree diameter, and their interaction. The MCMC chain for nitrogen fertilization is much noisier, and the chain struggles to converge.

```{r q2hide, results='hide'}
#----------------------------------------------------------------------------
## EXERCISE 2

# DATA
tmp    <- model.frame(y ~ j + i + diam*trt + nfert, data)
ij     <- apply( cbind(tmp$j, tmp$i), 1, paste0, collapse='-') # tree, plot
X      <- model.matrix(y ~ diam*trt + nfert, tmp)
y      <- tmp$y
n      <- length(y)

# JAGS
cat( "model{
    for(i in 1:n){
      y[i]  ~ dnorm(mu[i],tau)
      mu[i] <- intercept + diameter*X[i,2] + co2treat*X[i,3] + nfert*X[i,4] + inax*X[i,5]
    }

    tau     ~ dgamma(0.001,0.001)
    sigma   <- sqrt(1/tau)
    
    intercept ~ dnorm(0.0,1.0E-06)
    diameter ~ dnorm(0.0,1.0E-06)
    co2treat ~ dnorm(0.0,1.0E-06)
    nfert ~ dnorm(0.0,1.0E-06)
    inax ~ dnorm(0.0,1.0E-06)
  }
    ", fill=T, file="growthFixed.txt"
)
treeData   <- list(y = y, X = X, n = n )
variable.names <- c("intercept","diameter","co2treat","nfert","inax")
parNames <- c(variable.names,'sigma')
parInit <- function(){ list(sigma = runif(1,0,1)) }
growFit <- jags.model(data=treeData, file="growthFixed.txt")
update(growFit)
growFit <- coda.samples(growFit, variable.names=variable.names, n.iter=5000)

```

```{r q2plot}
tmp <- summary(growFit)
betaFixed <- cbind( tmp$statistics[,1:2], tmp$quantiles[,c('2.5%','97.5%')] )
print(betaFixed)
par(mar=c(3,3,1,1))
plot( as.mcmc(growFit) )
```
