---
title: "Homework 5 Responses"
author: "(Nicholas Bruns), Zeyi Han, Shubhi Sharma, Margaret Swift"
date: "2/24/2020"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=F}
#----------------------------------------------------------------------------
# LOADING
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, cache=TRUE)
pacman::p_load(ggplot2)
source('../clarkfunctions2020.R')
Rcpp::sourceCpp('../cppFns.cpp')
# any libraries and data here

#----------------------------------------------------------------------------
# FUNCTIONS
# put functions here

```


## Exercise 1 {#q1}
Using count data, I want to estimate a mean value.  Here is the Poisson likelihood:

$$
L(\mathbf{y};\lambda) = \prod_i^n Poi(y_i| \lambda) =
\prod_i^n \frac{\lambda^{y_i} e^{-\lambda}}{y_i!}  
\propto \lambda^{\sum_i y_i} e^{-n \lambda}
$$
The log likelihood is

$$
\log L \propto  n \bar{y}\log \lambda  -n \lambda
$$

I find the MLE by differentiating and solving for $\hat{\lambda}$:

$$
\frac{\partial{\log L}}{\partial \lambda} = \frac{n \bar{y}}{\lambda} - n 
$$
The solution is $\hat{\lambda} = \bar{y}$. Interpret this result.

```{r q1}
#----------------------------------------------------------------------------
## EXERCISE 1

```

## Exercise 2 {#q2}
Use Fisher Information to find the standard error of the mean of a normal sampling distribution.  Hint: you will need this:

$$
I = - \frac{d^2 \log L(\mu)}{d \mu^2}\bigg\rvert_{\hat{\mu}}
$$


```{r q2}
#----------------------------------------------------------------------------
## EXERCISE 2

```

## Exercise 3 {#q3}
Estimate the standard error for the exponential model using Fisher Information.  


```{r q3}
#----------------------------------------------------------------------------
## EXERCISE 3

```

## Exercise 4 {#q4}
For the cone example, I used the likelihood $Poi(y_i | \beta x_i)$.  Combine this likelihood with the prior $gamma(\beta | a, b)$ and answer the following:

a) What is the posterior density for $\beta$?

b) For simulated data sets of $n = 5$, how do the standard errors and credible intervals for this model compare with Fisher information?  

c) The form of the Bayesian standard error and the standard error from Fisher Information look different.  Can you explain why numerically they are similar? [Hint: think about sample size $n$].


```{r q4}
#----------------------------------------------------------------------------
## EXERCISE 4

```


## Appendix {#appendix}

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```


