---
title: "Homework N Responses"
author: "Zeyi Han, Shubhi Sharma, Margaret Swift"
date: '`r Sys.Date()`'
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=F}
#----------------------------------------------------------------------------
# LOADING
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, cache=FALSE)
pacman::p_load(ggplot2)
source('../clarkfunctions2020.R')
Rcpp::sourceCpp('../cppFns.cpp')
# any libraries and data here

#----------------------------------------------------------------------------
# FUNCTIONS
# put functions here

```


## Exercise 1 {#q1}

_Select a species to fit.  Use AIC as a model selection criterion to determine which combination of variables in `xdata` best fits the data._

```{r q1}
#----------------------------------------------------------------------------
## EXERCISE 1

```


## Exercise 2 {#q2}

_Do you believe that the MCMC converged to a posterior distribution?  If not, what might you do to convince yourself?_

```{r q2}
#----------------------------------------------------------------------------
## EXERCISE 2

```


## Exercise 3 {#q3}


_Divide up your group so that each determines how one of the following influences model selection.  Use DIC as your criterion._

_1. Different numbers of knots_

_2. Prior distributions on $\sigma^2$._

_3. Prior distributions on $\phi$._

```{r q3}
#----------------------------------------------------------------------------
## EXERCISE 3

```


## Exercise 4 {#q4}

_Interpret the contributions of the mean and the spatial random effects to predicted surfaces.  In other words, what do they mean?_

```{r q4}
#----------------------------------------------------------------------------
## EXERCISE 4

```


## Exercise 5 {#q5}

_Repeat exercise 3 with GJAM and compare results for the two models.  If there are differences, can you explain them?_

```{r q5}
#----------------------------------------------------------------------------
## EXERCISE 5

```

## Exercise 6 {#q6}

_Repeat the spatial analysis with a binomial model.  Use the `CARBayes` help page.  Interpret the analysis._

```{r q6}
#----------------------------------------------------------------------------
## EXERCISE 6

```

## Exercise 7 {#q7}
_Work through the lip cancer example in groups as if you were attempting to apply it to a data set of your own._

```{r q7}
#----------------------------------------------------------------------------
## EXERCISE 7

```
## Appendix {#appendix}

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```


