---
title: "Homework 7 Responses"
author: "(Nicholas Bruns), Zeyi Han, Shubhi Sharma, Margaret Swift"
date: '`r Sys.Date()`'
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=F}
#----------------------------------------------------------------------------
# LOADING
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, cache=TRUE)
pacman::p_load(ggplot2)
source('../clarkfunctions2020.R')
Rcpp::sourceCpp('../cppFns.cpp')
# any libraries and data here

#----------------------------------------------------------------------------
# FUNCTIONS
# put functions here

```


## Exercise 1 {#q1}

* Conduct simulation experiments to determine how different values of `sig` and 
`tau` affect the behavior of `x`.  

* Determine how these same parameters affect the standard deviation of `x` and 
`y`.

* Determine how the length of the series `nt` affects the standard deviation in 
`x` and in the differenced sequence of `x`.


```{r q1}
#----------------------------------------------------------------------------
## EXERCISE 1

```


## Exercise 2 {#q2}

Working in your group, analyze the following model in jags.  Start by simulating 
the data set.  Then write jags code to fit the model. 

$$x_t \sim N(\alpha + \beta x_{t-1}, \sigma^2)$$

$$y_t \sim Poi( exp(x_t) )$$


```{r q2}
#----------------------------------------------------------------------------
## EXERCISE 2

```


## Exercise 3 {#q3}

* Before running any more analysis, say what you think should be the variance on 
the process and the observations.  Justify your answer.

* How much 'weight' on the priors for $\sigma^2, \tau^2$ is needed to insure that 
the posterior distribution is close to your prior distribution?

* How does the prior on $\sigma^2, \tau^2$ affect estimates of $\mathbf{\beta}$?

* Repeat the analysis with DON as the response variable and interpret results.

* What is the answer to the motivating question: Can total dissolved nitrogen 
(TDN) and dissolved organic nitrogen (DON) be explained by temperature, 
precipitation, and chlorine concentation ([Cl]), and does it differs between 
coastal streams?

```{r q3}
#----------------------------------------------------------------------------
## EXERCISE 3

```



## Appendix {#appendix}

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```


