---
title: "Homework N Responses"
author: "(Nicholas Bruns), Zeyi Han, Shubhi Sharma, Margaret Swift"
date: '`r Sys.Date()`'
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=F}
#----------------------------------------------------------------------------
# LOADING
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, cache=FALSE)
pacman::p_load(ggplot2)
source('../clarkfunctions2020.R')
Rcpp::sourceCpp('../cppFns.cpp')
# any libraries and data here

#----------------------------------------------------------------------------
# FUNCTIONS
# put functions here

```


## Exercise 1 {#q1}

_ Conduct simlation experiments to determine how sample size and number of response variables affects estimates of the coefficients and covariance matrix._

```{r q1}
#----------------------------------------------------------------------------
## EXERCISE 1

```


## Exercise 2 {#q2}

_Jon has a survey with one binary response (`typeNames = 'PA'`) and three ordinal responses (ordinal counts, `'OC'`).  Simulate his data using `gjamSimData` and analyze it using `gjam`.  Using the documentation for `gjam`:

a) What are the partition estimates and what role do they play?

b) Which are better predicted by the fitted model, the `PA` data or the `OC` data?  Why might they differ?_

```{r q2}
#----------------------------------------------------------------------------
## EXERCISE 2

```


## Exercise 3 {#q3}

_Select 10 species from the BBS data and do a model selection (using DIC) to evaluate a limited number of variable combinations.

a) How do the DIC values compare with overall sensitivity?

b) With inverse prediction?_

```{r q3}
#----------------------------------------------------------------------------
## EXERCISE 3

```


## Exercise 4 {#q4}

Using the block of code that generates the predictive coefficient of variation, provide an interpretation of what you see.

```{r q4}
#----------------------------------------------------------------------------
## EXERCISE 4

```


## Appendix {#appendix}

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```


